---
title: "The Sixth Mass Extinction"
author: "Gi-Gi Lu, Chandler Sutherland"
maketitle: true 
output: github_document
---



```{r include=FALSE}
library("tidyverse")
library("httr")
library("jsonlite")
library("stringr")
knitr::opts_chunk$set(comment=NA)
```



## Mass Extinction: Are we experiencing the sixth great extinction? 

## Background
Species origination and extinction is a part of the natural process of the earth. In order to determine if the current rate of extinction on earth is abnormal, indicating a sixth mass extinction, we must first establish a background rate of extinction, and compare it to the current rate. We will be attempting to recreate a figure from [Ceballos et al (2015)](http://doi.org/10.1126/sciadv.1400253), using the IUCN Rest API. 


# Calculating Extinction Rates

First, we download a list of extinct species names and then build a function to extract the date they went extinct from the IUCN Rest API. 

```{r, message=FALSE}
extinct = read_csv("https://espm-157.github.io/extinction-module/extinct.csv")
extinct
```

```{r}
#Creates a list of apis to call 
api_call <- map2_chr(extinct$Genus, 
                     extinct$Species, 
                     function(genus, species)
                       paste0("http://api.iucnredlist.org/index/species/", genus, "-", species, ".json"))

#Get data from IUCN Rest API
resp <- map(api_call, 
            function(url){
              Sys.sleep(0.5)
              GET(url)
            })

status <- resp %>% map_int(httr::status_code)
good_resp <- resp[status == 200]
df <- map(good_resp, content, as = 'text') %>%
  map_dfr(fromJSON)

#Create dataframe from good response calls 
df_year <- df %>% select(genus, species, class, phylum, kingdom, rationale) %>% 
  mutate(year = as.numeric(str_extract(rationale, '\\d{4}'))) %>%
  filter(!is.na(year), !is.na(rationale)) %>%
  filter(year >= 1500, year <= 2015)

```


```{r}
hist(df_year$year, breaks = 20, xlab = 'Year', main = "Extinction Frequency per Year")
```

Now that we have a useable dataset, we can begin to investigate extinction rates. 

# Comparing Extinctions by Taxa

First, we computed the number of extinctions from 1500 - 1900 and from 1900 to present for a number of taxonomic groups. 

```{r}

k <- df_year %>% group_by(kingdom) %>% 
  filter(kingdom == "PLANTAE") %>% 
  summarize(early = sum(year <= 1900), late = sum(year > 1900)) %>% 
  rename("taxa" = kingdom) 

p <- df_year %>% group_by(phylum) %>% 
  filter(phylum == "CHORDATA") %>% 
  summarize(early = sum(year <= 1900), late = sum(year > 1900)) %>% 
  rename("taxa" = phylum) 

c <- df_year %>% group_by(class) %>% 
  filter(class == "MAMMALIA" | class == 'AVES' | class == 'AMPHIBIA' | class == 'REPTILIA' | class == 'INSECTA') %>%
  summarize(early = sum(year <= 1900), late = sum(year > 1900)) %>% 
  rename("taxa" = class) 

early_late <- bind_rows(k, p, c)
early_late

```

> Relative to Table 1 of [Ceballos et al (2015)](http://doi.org/10.1126/sciadv.1400253), we found fewer extinctions per taxa. This is probably due to our method of extracting the year, since we only look for some matching year in "rationale" and there is no documentation specifying that this is the year they went extinct. However, our numbers are similar proportionally.


## Weighing by number of species

In order to have a meaningful comparison of extinction rates, we must weight our number of species going extinct at each time point by the total number of species in each taxonomic group. That is, we will compute the number of extinctions per million species per year (MSY; equivalently, the number extinctions per 10,000 species per 100 years). 

First, we compute how many species are present in each of the taxonomic groups by creating a table that includes all assessed species by quering the IUCN API again. 


```{r, message = FALSE}
#Gets the count of all the species
getCount <- function() {
  resp <- GET("http://apiv3.iucnredlist.org/api/v3/speciescount?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")
  status <- http_error(resp)
  if (!status) {
     out <- content(resp, as = "text")
    df <- jsonlite::fromJSON(out)
    return(as.numeric(df$count))
  }
 
}
speciesCount <- getCount()
 
#Creates a list of API calls to species information 
api_species <- map(seq(0, ceiling(speciesCount/10000)-1), function(i)
  GET(paste0( "http://apiv3.iucnredlist.org/api/v3/species/page/", as.character(i),"?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee"))
  )
spec <- map(api_species, content, as = 'text') %>%
  map_dfr(function(resp) {
    jsonlite::fromJSON(resp)[["result"]]
    })

#create a summary table by taxa, counting the number of species to get a species_count
k2 <- spec %>%
  filter(kingdom_name == "PLANTAE") %>% 
  group_by(kingdom_name) %>%
  summarize(species_count = n()) %>%
  rename("taxa" = kingdom_name) 

p2 <- spec %>% 
  filter(phylum_name == "CHORDATA") %>% 
  group_by(phylum_name) %>% 
  summarize(species_count = n()) %>%
  rename("taxa" = phylum_name) 

c2 <- spec %>%
  filter(class_name == "MAMMALIA" | class_name == 'AVES' | 
           class_name == 'AMPHIBIA' | class_name == 'REPTILIA' | 
           class_name == 'INSECTA') %>%
  group_by(class_name) %>%
  summarize(species_count = n()) %>%
  rename("taxa" = class_name) 


all <- bind_rows(k2, p2, c2)

#calculate million species-year for each taxa 
msy <- all %>% right_join(early_late) %>%
  mutate(extinct = early + late) %>% 
  mutate(MSY = extinct/species_count*100) %>%
  select(taxa, MSY)

msy
```

Our MSY is lower than the proposed baseline of 2 MSY. Howver, this is most likely due to issues with our algorithm in how we extract dates from the rationale. We are undercounting the number of extinct species severely using this method. 


## Improving our algorithm

Our method of extracting date from rationale is flawed because we just pull the first four digit number in the rationale. However, without the context of the sentence this could be the incorrect year. Oftentimes the first year listed is the last time it was seen in the wild, not the date declared extinct. We also encountered missing values because the data was not meant to be parsed in this way, ie to extract extinction year from a text block.  

